### 2020-02-01 Common/chapters/paper_reproduction.tex
Julien and Johan: data processing for "paper reproduction" and "transfer learning" chapters

\section{Process overview}
Schematically, the entire experiment process can be represented as shown in figure \ref{fig:paper_reproduction_process}. The PROSTATEx dataset is composed of samples whose clinical significance is provided (labeled data) and samples without this information (unlabeled data). The latter are used for the challenge, which consists in predicting the probability of patients lesions to be malignant (class 1) or benign (class 0). In both cases, the data is processed before been used (see \ref{prostatex_data_processing} for details). After the processing, the labeled data are split into training, validation and test sets using respectively 80\%, 10\% and 10\% of all available data for each set. The training set is then given in batches as input to the neural network, which update its weight consequently. At the end of each epoch, the current model is tested on the validation set (that the model has never seen previously) and the metrics are plotted on Tensorboard (see \ref{paper_tensorboard}). If the current model has a bigger AUC and a bigger accuracy than the previous best model, it becomes the best model and it is saved.  When the training phase reaches the defined number of epochs, the model is then tested on the test set. This same model is also used to take part in the PROSTATEx challenge to output predictions into a .csv file for each patient and findings.

@@

\section{PROSTATEx: Data processing}
\label{prostatex_data_processing}
\subsection{Dataset description}
\label{prostatex_dataset_description}
\subsection{From DICOM to NumPy arrays}
\subsection{From NumPy arrays to augmented stacked images}
\subsection{From NumPy arrays to augmented non-stacked images}
\subsection{Data processing verification}
\subsubsection{Cropping verification using red dots}
\subsubsection{Alignment visualization}

>>>>

\section{Process overview}
Schematically, the entire experiment process can be represented as shown in figure \ref{fig:paper_reproduction_process}. The PROSTATEx dataset is composed of samples whose clinical significance is provided (labeled data) and samples without this information (unlabeled data). The latter are used for the challenge, which consists in predicting the probability of patients lesions to be malignant (class 1) or benign (class 0). In both cases, the data is processed before been used (see \ref{sec:prostatex_data_processing} for details). After the processing, the labeled data are split into training, validation and test sets using respectively 80\%, 10\% and 10\% of all available data for each set. The training set is then given in batches as input to the neural network, which update its weight consequently. At the end of each epoch, the current model is tested on the validation set (that the model has never seen previously) and the metrics are plotted on Tensorboard (see \ref{paper_tensorboard}). If the current model has a bigger AUC and a bigger accuracy than the previous best model, it becomes the best model and it is saved.  When the training phase reaches the defined number of epochs, the model is then tested on the test set. This same model is also used to take part in the PROSTATEx challenge to output predictions into a .csv file for each patient and findings.

@@

\section{PROSTATEx: Data processing}
\label{sec:prostatex_data_processing}
\subsection{Dataset description}
\label{sec:prostatex_dataset_description}
The SPIE-AAPM-NCI PROSTATEx Challenge dataset is publicly available on the Cancer Imaging Archive. This dataset is composed of multiparametric MRIs (T2W, DWI, ADC, DCE, PD, Ktrans) for a total of 204 training patients and 208 challenge patients. Images were taken under the sagittal, transverse and coronal planes.

The PROSTATEx dataset comes with two CSV files for the training set. The first one, \textit{ProstateX-Findings-Train.csv}, lists all findings with their clinical significance. Multiple findings can be associated with the same patient (ProxID). The second one, \textit{ProstateX-Images-Train.csv}, gives information about where to find the right DICOM file for each patient and each finding. Important labels are "ProxID" (patient ID), "fid" (finding ID $\in [1,\infty]$), "ClinSig" (clinical significance, TRUE or FALSE), "DCMSerNumber" (digit before the dash in the folder name containing DICOM files), "ijk" (position of the lesion: slice number $k$ at coordinates $(i,j)$, $i,j,k \in [0,\infty]$) and "VoxelSpacing" (3-dimensional vector representing the correspondance between a pixel and the space it occupies in the real world). Both CSV files are complementary to each other. Regarding the challenge patients, two analog CSV files are provided. The only difference is the absence of clinical significance. Finally, images are in the DICOM file format (see section \ref{sec:DICOM}).

\subsection{Methodology}
The methodology described in this section is the closest replication possible of the authors' data processing methodology. Most of it were reproductible in a similar way, apart from the manual lesion contouring which was performed by a qualified radiologist in the authors' case. Furthermore, the authors didn't mention any participation in the official PROSTATEx challenge. In order to get an unbiased evaluation of the performance of our model, these images were also processed as explained below in order to take part in the challenge, but were not augmented. Nevertheless, the authors created their own test set by splitting the PROSTATEx training images into a training set ($80\%$), a validation set ($10\%$) and a test set ($10\%$). They mentioned that the test set was augmented $11$ times but gave no information about the training and validation sets. Since the training set is imbalanced (3 false for 1 true), the true class was augmented more times than the false class ($60x$) in order to create balanced training and validation sets.

\subsection{From DICOM to NumPy arrays}
\label{sec:DICOMtoNumPy}
Before anything else, T2W, DWI and ADC grayscale images were used, which means that DCE, PD and Ktrans images were left aside. Moreover, only images showing the prostate under the transverse plane were used. The reason for this is that the tumors are much more visible under this perspective. 
The first step consisted in converting DICOM files. Algorithm \ref{alg:PROSTATEx_preprocessing} describes the steps involved in converting PROSTATEx's DICOM files to NumPy arrays. The right slices were found thanks to the two CSV files. Important information such as the patient ID, the sequence, the lesion location and the voxel spacing were used as file output names, which allows to use these files independently for the next steps (i.e. without relying on the CSV files).

\begin{algorithm}
    \caption{PROSTATEx preprocessing}
    \label{alg:PROSTATEx_preprocessing}
    \begin{algorithmic}[1] % The number tells where the line numbering should start
        \Procedure{main}{$dataset\_folder, findings\_CSV, slices\_CSV, output\_folder$}
        		\State Create output directories: $"output\_folder/True", "output\_folder/False"$\\
        		\State $findings \gets read\_CSV(findings\_CSV)$ \Comment{ProstateX-Findings-Train.csv}
        		\State $slices \gets read\_CSV(slices\_CSV)$ \Comment{ProstateX-Images-Train.csv}
            \State $meta \gets merge(findings, slices)$\Comment{Both CSV files are complementary to each other.}\\
            \For{$row$ in $meta$}
            		\State $patient\_id \gets row["ProxID"]$
            		\State $finding\_id \gets row["fid"]$
            		\State $mri\_type\_number \gets row["DCMSerNumber"]$
            		\State $clinical\_significance \gets row["ClinSig"]$
            		\State $img\_i, img\_j, img\_k \gets row["ijk"]$
            		\State $slice\_number \gets img\_k + 1$ \Comment{CSV indexing in $[0,\infty]$, DICOM in $[1,\infty]$}\\
                \For{$visit\ in\ patient\_id$'s folder}
                		\For{$mri\_type$ in $visit$}
                			\If {$mri\_type$ starts with "$mri\_type\_number$-"}
                				\For{$dicom\_file$ in $mri\_type$}
                					\If {$slice\_number == dicom\_file.InstanceNumber$}
                						\State{$slice \gets normalize\_dicom(dicom\_file)$} \Comment{see section \ref{sec:dicom_data_manipulation}}
                						\State{Save $slice$ in $"output\_folder/clinical\_significance"$}
                					\EndIf
                				\EndFor
                			\EndIf
            			\EndFor
            		\EndFor
            \EndFor
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

\subsection{From NumPy arrays to augmented stacked images}
\label{sec:numpyToAugmentedStacked}
This step can be split into two substeps: aligning and stacking images before augmenting them. Once full images were converted to NumPy arrays, the images related to a specific lesion of a specific patient needed to be aligned and resampled to the same resolution. In fact, T2-weighted images have a much higher resolution than the DWI and ADC images (in this dataset at least). Without this operation, a single pixel on a DWI image would have covered a lot more tissue than a pixel of the corresponding T2-image. To perform the alignment, the lesion was localized on the three sequences than to the CSV file. Then, the goal was to crop a large patch which contained the same amount of tissue on the three images. This patch had to be centered on the lesion to ease the augmentation process. So, a fixed patch size was defined for the T2-image since its resolution was the highest. Then, using the voxel spacing information, the patch size required to cover the same amount of tissue was computed for the two other sequences (DWI, ADC). Finally, the three images were cropped, resized to the same resolution and stacked. Stacking images consists in putting each single image (grayscale) into a single array, as three different channels. At this point, the pixel $i,j$ of the three channels represents the exact same tissue, which means that the lesion is at the same position and is covered by the same number of pixels on each channel. This approach increases the probability of detecting a cancer by ensuring a good visibilty for each lesion, since the latter is not necesarily as visible on the three images.

An important part of this process is the normalization. In fact, images were normalized before beeing stacked, based on a Z-score, i.e.

\begin{equation}
\label{eq:normalization}
	Pixel_{{i,j}_{normalized}} = \frac{Pixel_{i,j} - \mu}{\sigma}
\end{equation}

\noindent where $Pixel_{i,j} \in [0,255]$ is a grayscale pixel value, $\mu$ is the mean value of all images of the corresponding sequence for this patient and $\sigma$ the standard deviation of all images of the corresponding sequence for this patient. According to the authors, normalizing each sequence of each patient separately allows to keep slight contrast nuances which ultimately help the final diagnosis. In other words, all T2-weighted images belonging to a specific patient were concatenated into a single NumPy array. Then, the mean and standard devation of the array were computed. These values served as $\mu$ and $\sigma$ to normalize the T2-images of this specific patient. Same for this patient's DWI and ADC images. The same process was applied to all patients separately.

Finally, the dataset was augmented in the same way as the authors. Concretely, images were rotated ($-20$ to $20^\circ$), flipped horizontally (probability of $0.5$) and slid horizontally (value $\in [-1, 0, 1]$ pixel). These techniques alone allowed to create a sufficient amount of data. Therefore, two other augmentation methods used by the authors (horizontal stretching by a factor $\in [0.9,1.1]$ and vertical sliding) were not made use of. Images were exported as NumPy arrays.

\subsection{From NumPy arrays to augmented non-stacked images}
Another way of processing and using the data was experimented with. Instead of stacking the images, each image was cropped and augmented separately. At the end, each image was given as separate input to the neural network.

\subsection{Data processing verification}
\label{sec:dataProcessingVerification}

\subsubsection{Cropping verification using red dots}
In order to check the correctness of the augmentation process, images with a red dot at the lesion coordinates were generated. First of all, a red dot was placed on the full image thanks to the "i,j,k" attribute in the CSV file. This full image was exported. Then, images were augmented in the exact same manner as described in section \ref{sec:numpyToAugmentedStacked}. The expected result satisfies the following properties:
\begin{itemize}
	\item The red dot is localized in the center of the cropped and augmented image.
	\item The three cropped and augmented images within a stacked image contain the exact same tissue. 
	\item The three cropped and augmented images within a stacked image are not rotated/shifted/flipped differently. 
	\item A lesion is visible to the naked eye. 
\end{itemize}

\noindent Figure \ref{fig:reddot} shows an example generated during this test. While beeing exactly the same for the 3 images, the red dot looks bigger on the images whose resolution was smaller since it was drawn before resampling the images. These images come from the same exam performed on the same patient on the same day.

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth, keepaspectratio=true]{./figures/test_red_dot.png}
\caption{Red dot test - Patient 0082, Finding ID 1\\From left to right: T2, DWI, ADC.}
\label{fig:reddot}
\end{figure}

\subsubsection{Alignment verification}
In addition to the previous test, an additional one aiming at checking the alignment of the images was created. This test ensures that the three images show the exact same content at the same spot over the three channels.

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth, keepaspectratio=true]{./figures/alignment.png}
\caption{Alignment - Patient 0242 - Finding ID 1}
\label{fig:alignment}
\end{figure}

### 2020-02-01 Common/chapters/transfer_learning.tex
Johan and Julien: paper reproduction chapter started
Julien and Johan: data processing for "paper reproduction" and "transfer learning" chapters

\section{Data processing}
\subsection{PROSTATEx}
\subsubsection{From DICOM to augmented DWI NumPy arrays}
Keep only the DWI because white lesion, crop the lesion etc

\subsection{LungCTChallenge}
\subsubsection{Dataset description}
\subsubsection{From DICOM to augmented NumPy arrays}

\subsection{Kaggle Brain}
\subsubsection{Dataset description}
\subsubsection{Ground truth creation}
\subsubsection{From PNG to NumPy arrays}
Manually create the csv file from scratch, crop the lesion etc

\subsection{Verification}
\subsubsection{Visual checking}

>>>>

\section{Data processing}
\subsection{PROSTATEx}
As the other datasets used in this work don't contain multi-sequence exams (single-sequence MRIs and CT scans), it was required to extract single channel images from the PROSTATEx dataset. DWI images were selected because lesions are white on this sequence, as well as on CT scans. Hence, choosing DWI images ensured a consistent representation of tumors from one dataset to the other, which helps the cancer detection.

The overall process is the same as the one described in section \ref{sec:numpyToAugmentedStacked}, apart from the alignment and the stacking. A T2-weighted image was still used to align, crop and resize the corresponding DWI image but was not exported at the end, resulting in a single channel image (NumPy array).

Ideally, transfer learning should rely on multi-sequence MRIs only. In fact, the extra information provided by different sequences increases the performance a lot. However, publicly available datasets either don't contain the clinical significance of the samples or were purely made for segmentation purposes (where labels are also images).

\subsection{LungCTChallenge}
\subsubsection{Dataset description}
Lung CT Challenge is composed of two different subdatasets of CT scans: one is called "calibration set" (10 patients) and the other "test set" (60 patients). Each patient can have 1 or multiple findings. Since labels were provided for both sets and the amount of data is fairly low, they were merged and used as a global training set. For each finding, an abnormal mass is visible in the lungs, which means that the classification is going to differentiate malignant from benign tumors.

Regarding labelling, two Excel files, \textit{TestSet\_NoduleData\_PublicRelease\_wTruth} and \textit{CalibrationSet\_NoduleData}, contain labels for these images. In order to facilitate the handling of information in the code, two CSV files were manually created: \mbox{\textit{TestSet.csv}} and \textit{CalibrationSet.csv}.
Contrary to PROSTATEx, more than two labels were used in this dataset. Both "malignant" and "Primary lung cancer" were considered as positive, whereas "benign" and "Benign nodule" were treated as negative. A third label called "Suspicious malignant nodule" appeared two times. Since the diagnosis was not clearly defined for those images, they were not included in the training data to avoid any noise.

\subsubsection{From DICOM to augmented NumPy arrays}
As PROSTATEx, the processing was split into two parts: preprocessing (DICOM to NumPy arrays) and augmentation (NumPy arrays to cropped augmented NumPy arrays.) Algorithm \ref{alg:LungCTChallenge_preprocessing} shows the various preprocessing steps.

\begin{algorithm}
    \caption{Lung CT Challenge preprocessing}
    \label{alg:LungCTChallenge_preprocessing}
    \begin{algorithmic}[1] % The number tells where the line numbering should start
        \Procedure{main}{$dataset\_folder, train\_CSV, test\_CSV, output\_folder$}
        		\State Create output directories: $"output\_folder/True", "output\_folder/False"$\\
        		\State $csv\_training \gets read\_CSV(train\_CSV)$ \Comment{CalibrationSet.csv}
        		\State $csv\_test \gets read\_CSV(test\_CSV)$ \Comment{TestSet.csv}
            \State $csv\_concatenated \gets concat(csv\_training, csv\_test)$\Comment{Both CSV files contain similar information about different patients.}\\
            \For{$row$ in $csv\_concatenated$}
            		\State $patient\_id \gets row["Scan\ Number"]$
            		\State $slice\_number \gets row["Nodule\ Center\ Image"]$ \Comment{Value in $[1,\infty]$}
            		\State $finding\_id \gets row["Nodule\ Number"]$
            		\State $clinical\_significance \gets row["Diagnosis"]$\\
            		\If{$clinical\_significance$ == $"malignant"$ or $"Primary\ lung\ cancer"$}
            			\State{$clinical\_significance \gets True$}
            		\ElsIf{$clinical\_significance$ == $"benign"$ or $"Benign\ nodule"$}
            			\State{$clinical\_significance \gets False$}
            		\ElsIf {$clinical\_significance$ == $"Suspicious\ malignant\ nodule"$}
            			\State{Continue}
            		\EndIf\\
                \For{$visit\ in\ patient\_id$'s folder}
                		\For{$mri\_type$ in $visit$}
                			\For{$dicom\_file$ in $mri\_type$}
                				\If {$slice\_number == dicom\_file.InstanceNumber$}
                					\State{$slice \gets normalize\_dicom(dicom\_file)$} \Comment{see section \ref{sec:dicom_data_manipulation}}
                					\State{Save $slice$ in $"output\_folder/clinical\_significance"$}
                				\EndIf
                			\EndFor
            			\EndFor
            		\EndFor
            \EndFor
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

The augmentation process is close to the one applied to PROSTATEx. An advantage Lung CT Challenge has it that all images have the same resolution (512x512 pixels), making the alignment process used for PROSTATEx useless. Nevertheless, lung tumors are particular. What differentiates them from the others is their look. In fact, lung tumors usually look like spider webs with a central nodule and branches coming out of it, whereas brain and prostate tumors tend to have well-defined boundaries. Therefore, a larger patch size was chosen in order to keep the extra information provided by the branches in the peripheral tissue. Furthermore, a perfect cropping method would only crop the interesting part of the tissue and a margin around it. However, tumor size varies a lot. It can either be extremely small (a dot) or large (massive nodule with long branches). Consequently, the selected patch size makes the largest tumors fit perfectly in the cropped area, inducing more tissue for smaller tumors. Naturally, there were few outliers whose size was a lot larger than the chosen patch size. If the patch size was set accordingly, entire lungs were sometimes visible in other patients' case. For this reason, these large tumors were treated as outliers. Lastly, and contrary to PROSTATEx, the dataset was more or less balanced (39 positives and 42 negatives). Therefore, each class was augmented the same amount of times (240x).

Images were normalized with respect to equation \ref{eq:normalization}. The mean and standard deviation were computed over the whole dataset.

\subsection{Kaggle Brain}
\subsubsection{Dataset description}
This dataset is publicly available on Kaggle. All images come from single-sequence MRIs. It is composed of two folders: one containing images with cancerous tumors and another one containing images either without tumor or with a non-cancerous tumor. Images whose quality was extremely low (resolution, picture of a screen, artefacts, etc.) were manually removed. In other cases, the image quality was good but disturbing objects such as left/right markers, arrows pointing the tumor, etc. were visible. These objects were manually removed using the object removal tool in Photoshop. File names and file formats were heterogeneous.

\subsubsection{Ground truth creation}
\label{sec:braingroundtruth}
The particularity of this dataset is that no information apart from the clinical significance is provided. Therefore, it was necessary to create a ground truth CSV file from scratch so that the same processing methods as the other datasets could be applied on this one.

The first step was to standardize the file names and formats. The chosen convention is "$[yes|no]{index}.jpg$". The vast majority of the files were in the JPG file format, so the few PNG images were converted. Then, a CSV file containing the following columns was created: "PatientID", "Nodule Center Position", "fid", "Diagnosis". The "PatientID" is the filename without extension that was set during the first step (for example "$yes99$" or "$no3$"). The "Nodule Center Position" is the pixel coordinate of the lesion. Fortunately, brain tumors are clearly visible on brain MRIs, which eased the process. To find the coordinates, every single image was opened at its original size in an image visualization software. A rectangle selection tool was used to draw a rectangle from the top left corner of the image (because the top left pixel corresponds to the pixel at the coordinates $(0,0)$ in NumPy arrays) to the center of the lesion. The pixel width and height of this rectangle gave the "i,j" coordinates of the lesion. These coordinates were reported manually. Then, the "fid" corresponds to the Finding ID. This parameter would be more pertinent if the same "PatientID" could have multiple findings, i.e. if the same patient could have multiple tumors. Since this information is made use of during the image processing of the other datasets, it was also added here. In this particular case, the "fid" was always set to "1". Finally, the "Diagnosis" was defined as "False" or "True" according to the folder which the image was in.

\subsubsection{From JPG to NumPy arrays}
Images had to be converted to single-channel NumPy arrays (grayscale) so that the values could be normalized further on. First, a mere conversion involving no content modification apart from going from three to one channel was performed. Second, images were normalized according to equation \ref{eq:normalization}. The mean and standard deviation were computed over the whole dataset. Both classes were augmented by a common factor (200x) since both class contain a relatively close number number of instances (133 positives and 96 negatives). Similarly to Lung CT Challenge, tumor sizes vary substantially. The same methodology was applied regarding the patch size, i.e. choosing a size which fits the majority of large tumors perfectly. Even though the images all show the head under the same perspective (same height/width ratio, visible skull, black margin around the skull), images were probably collected from different sources, which explains the heterogeneity of resolutions. As a consequence, the patch size could not be a fixed number of pixels for all images because the cropped area would be the entire head on one image and a small portion of the tumor on another. The trick consisted in finding the horizontal and vertical portions of the image occupied by the tumor. In fact, cropping an area centered on the lesion whose height was $\frac{2}{7}$ of the image and whose width was $\frac{1}{4}$ gave good results.

\subsection{Verification}
\subsubsection{Visual checking}
Regarding the Kaggle brain dataset, the data generated was paid close attention to since the ground truth CSV file was created by hand (see section \ref{sec:braingroundtruth}). In fact, each generated image was taken a close look at, which allowed to spot 3 mistakes over the whole dataset. They concerned typos in the "Nodule Center Position", which created pitch black images since the cropped area was outside.
Like PROSTATEx (see section \ref{sec:dataProcessingVerification}), control images with red dots were generated for both datasets.
